## Data

# The data comes from a university registrar's office. The code book for the variables are available in the file code-book.txt

# Upload the drop-out.csv data into R as a data frame. 
```{r}
DO <- read.csv("drop-out.csv")
```

# Randomly select 25% of the students to be the test data set and leave the remaining 75% for your training data set.
```{r}
library(dplyr)
set.seed(1234)
DOUnique <- as.data.frame(unique(DO$student_id))

#Samples 25% of the Student IDs
DO25 <- DOUnique %>% sample_frac(0.25)
names(DO25) <- c("student_id")

#We Use Inner Join to Take Just the Students with Students IDs in the 25% Sample Data Frame
#We Use Anti Join to Take the Students with Student IDs NOT in the Sample Data Frame (remaining 75%)
training <- inner_join(DO, DO25, by = "student_id")
testing <- anti_join(DO, DO25, by = "student_id")


```

#Complete and Years have a significant relationship. We see that years is strongly correlated with other variables as well. Most people who completed the course were in their first year or early on in their years. We project that years will be an important split in our decision tree. 
```{r}
#Set Color for Scatterplot Matrix, We Correlate The Variables Against The Yes and No Outcomes Of Completion
cols <- character(dim(DO)[1])
cols[which(DO$complete == "yes")] <- "yellow3"
cols[which(DO$complete == "no")] <- "slateblue4"

#Scatterplot Matrix
pairs( ~ years + entrance_test_score + enroll_date_time + courses_taken + online + course_id, data = DO, col = cols)

```

```{r}
library(caret)
install.packages("rpart.plot")
library(rpart.plot)

#Remove the student_id variable that we do not want to use in the model
training2 <- training[,c(2:10)]

#Define the control elements we would like to use
ctrl <- trainControl(method = "repeatedcv", #Tell caret to perform 10-fold cross validation
                repeats = 3, #Tell caret to repeat each fold three times
                classProbs = TRUE, #Calculate class probabilities for ROC calculation
                summaryFunction = twoClassSummary)

#Define the model
cartFit <- train(complete ~ ., #Define which variable to predict
                data = training2, #Define the data set to train the model on
                trControl = ctrl, #Tell caret the control elements
                method = "rpart", #Define the model type
                metric = "ROC") #Tell caret to calculate the ROC curve
                 #Center and scale the data to minimize the 

#Check the results
cartFit
                
#Plot ROC against complexity 
plot(cartFit)

#Plot the Tree
prp(cartFit$finalModel, box.palette = "Reds", tweak = 1.2)
```

#When running the tree multiple times, we see that years, course_taken, and course_id are consistent nodes (top branches). We can conclude that these are among the most significant. Years represents how long the students have been in their program. We see that those in their second, third, and so on years (years greater than 1) were more likely to drop out of their courses. This is consistent with the scatterplot matrix we've generated above which shows that those in their first year were more willing to stay to complete the class. Within the the students who were in their first years, those who have taken MORE than four courses were more likely to complete the courses. Notably, the courses_taken decision rule goes against the intuition provided by the first node which suggests that the longer a student is in the program, the more likely they would drop out out of a program. The course_id decision rule indicates that students who take courses with higher course_ids are more likely to complete. This might indicate that these courses (higher id courses) are major requirements and so students are less likely to drop.


#The plot displays ROC regressed against CP (Complexity Parameter). The complexity parameter is the threshold associated to comparing the cost of adding an additional node. The ROC curve is a tool to see if predictive models can distinguish between true positives and negatives; this is done by regressing sensitivity (probability of accurately predicing a real positive is positive) against 1 - specificity (probability of predicing a real negative will be positive). The ROC value is the area under this curve and depicts accuracy. The plot displays a negative linear relationship. Thus, as the cost to split more nodes in the decision tree increases, the accuracy of the model generally decreases. 


#Running the model on the test data we see that it has an accuracy of roughly 90%. This is a very high accuracy rate. Through this metric alone, we can conclude that the it is a successful model for predicting whether a student will drop out of the class or not. 

```{r}
testing2 <- testing[,c(2:10)] #Remove the student_id variable that we do not want to use in the model

#Generate prediction using previously trained model
cartClasses <- predict(cartFit, newdata = testing2)

#Generate model statistics
confusionMatrix(data = cartClasses, testing2$complete)

```

```{r}
install.packages("party")
library(party)
ctreeFit <- train(complete ~ .,
                data = testing2,
                trControl = ctrl,
                method = "ctree",
                metric = "ROC")

ctreeFit

plot(ctreeFit)
print(ctreeFit)

#We Plot the Tree Model for the Conditional Inference Tree
plot(ctreeFit$finalModel)
plot(ctreeFit$finalModel, type = "simple")
```
#The important model attriutes remains the same: years and course_taken remain the significant attributes. The model is crowded but obtains an accuracy high into the 80%. Thus, the model is successful in predicting student drop out.

#The plot displays the ROC value (area under ROC curve) regressed agaist the 1 - P-value threshold. The conditional inference tree uses significance test procedures to select variables of interest. The plot displays a mountain-type, pointed shape. The peak describes the value where the model is most accurate. Reading from the graph, we see that a 1 - P-value Threshold of 0.5 produces the highest model accuracy. 

```{r}
TESTCI <- testing[,c(2:10)] #Remove the student_id variable that we do not want to use in the model

#Generate prediction using previously trained model
ctreeFitClass <- predict(ctreeFit, newdata = TESTCI)

#Generate model statistics
confusionMatrix(data = ctreeFitClass, TESTCI$complete)

```
```{r}
install.packages("C50")
library(C50)
c50Fit <- train(complete ~ .,
                data = training2,
                trControl = ctrl,
                method = "C5.0",
                metric = "ROC",
                preProc = c("center", "scale"))

c50Fit

plot(c50Fit)

c50Classes <- predict(c50Fit, newdata = testing2)

confusionMatrix(data = c50Classes, testing2$complete)
```

## Compare the models
```{r}
resamps <- resamples(list(cart = cartFit, ctree = ctreeFit, cfiveo = c50Fit))
summary(resamps)
```

#The model summary tells us the quartiles of the models in terms of its ROC, Sensitivity, and Specificity. We that the C5 package generates the highest mean and medians for ROC and has the highest accuracy.

#My chosen model is C5. The important variables are years, courses_taken, and course_id. While this gives us some insight into why students are dropping, converting them into actions to reduce dropout rate may be difficult as it doens't identify those who were dropping out because they were struggling. Looking into a smaller subset of particular classes with other variables may give us better information.  